# Theme: Carbon and Water Fluxes

**What:** Net ecosystem exchange of CO2 and evapotranspiration in terrestrial ecosystems

**Where**: 10 NEON sites across the conterminous U.S. The images below show the flux tower at the center of each site.

**When**: Daily forecasts for 35-days in the future. Forecast submissions are accepted daily throughout 2022.  The only requirement is that submissions are predictions of the future at the time the forecast is submitted.

**Why:** Carbon and water cycling are fundamental for climate and water regulation services provided by ecosystems

**Who**: Open to any individual or team that registers

**How**: [REGISTER](https://nd.qualtrics.com/jfe/form/SV_9MJ29y2xNrBOjqZ){target="_blank"} your team and submit forecast. If you registered for the Round 1 (2021) and are using the same team and method then you do not need to re-register.

The video below is an overview of the Terrestrial Carbon and Water Fluxes Challenge that was recorded for the [2021 Early Career Annual Meeting](https://ecoforecast.org/ecological-forecasting-early-career-annual-meeting/){target="_blank"}

<iframe width="560" height="315" src="https://www.youtube.com/embed/WmFsACm-TMk" title="YouTube video player" frameborder="0" allow="accelerometer; autoplay; clipboard-write; encrypted-media; gyroscope; picture-in-picture" allowfullscreen></iframe>

We held a Q&A session on January 22, 2021. You can find a recording from that session [HERE](https://www.youtube.com/watch?v=3rOVlAX9YH8&feature=youtu.be){target="_blank"}.

```{r echo = FALSE, message = FALSE}
library("tidyverse")
```

## Overview

The exchange of water and carbon dioxide between the atmosphere and the land is akin to earth’s terrestrial ecosystems breathing rate and lung capacity. The water available to plant roots plays a critical role in plant function, and subsequently represents a predominant source of uncertainty for predictions of how much carbon is entering or exiting an ecosystem. One of the best ways to monitor changes in the amount of carbon and water in an ecosystem is the eddy-covariance method. This method observes the net amount of carbon and water entering and exiting ecosystems at half-hourly timesteps, which is important because it can provide information on ecosystem processes such as photosynthesis, respiration, and transpiration, their sensitivities to ongoing climate and land use change, and greenhouse gas budgets for carbon accounting and natural climate solutions. Forecasts of carbon uptake and release, water use, and soil moisture can provide insights into future production of food, fiber, timber, and carbon credits. Additionally, forecasts will highlight the influence that stress and disturbance have on carbon and water cycling.

## Challenge 

This design challenge asks teams to produce forecasts of net ecosystem exchange of carbon dioxide (NEE) and latent heat flux of evapotranspiration (LE) across four NEON sites with differing climates. These target variables are important because they can be used to inform energy budgets and further reduce uncertainty in the CO2 sink or source behavior of the terrestrial biosphere. 
This forecasting challenge asks teams to forecast NEE, LE, and soil moisture at either the 30-minute or daily time step over the next 35-days using NOAA Global Ensemble Forecast System weather forecasts as drivers (if forecasting model uses meteorological inputs).  Forecasts can be submitted each day in 2022. 

The challenge will take place using the eddy covariance flux towers at 12 NEON sites.

Users are asked to submit their forecast of measured NEON NEE and LE, along with uncertainty estimates and metadata. Any existing NEE and LE may be used to build and improve the models used to generate forecasts. Other data can be used so long as they are not from the month being forecast and the data are made publicly available (minimum of URL, but ideally a script) and accessible to all teams in the challenge.

## Data: Targets 

The challenge uses the following NEON data products:  

[DP4.00200.001](https://data.neonscience.org/data-products/DP4.00200.001){target="_blank"}: Bundled data products - eddy covariance   

A file with previously released NEON data that has been processed into “targets” is provided below. The same processing will be applied to new data that are used for forecast evaluation. A processing script is available on the neon4cast-terrestrial GitHub repository: https://github.com/eco4cast/neon4cast-terrestrial/blob/master/02_terrestrial_targets.R

### Net ecosystem exchange

**Definition**

Net ecosystem exchange (NEE) is the net movement of carbon dioxide from the atmosphere to the ecosystem.  At the 30-minute time resolution is reported as $\mu$mol CO<sub>2</sub> m<sup>-2</sup> s<sup>-1</sup>.  At the daily time resolution it is reported as g C m<sup>-2</sup> day<sup>-1</sup>. Negative values correspond to an ecosystem absorbing CO<sup>2</sup> from the atmosphere, positive values correspond to an ecosystem emitting CO<sub>2</sub> to the atmosphere.

**Motivation**

NEE quantifies the net exchange of CO<sub>2</sub> between the ecosystem and the atmosphere over that 30-minute or daily time period.  Assessing skill at predicting 1/2 hourly - sub daily measurements provides more insight into ability to capture diel processes. The diel curve contains information on how plants and soil immediately respond to variations in meteorology. 

Making daily predictions will allow us to rapidly assess skill and provide information in a timeframe pertinent to inform and implement natural resource management. It also allows for models that do not produce sub-daily estimates to participate

### Latent heat flux

**Definition**

Latent heat flux is the movement of water as water vapor from the ecosystem to the atmosphere.  It is reported as W m<sup>-2</sup> (equivalent to J m<sup>-2</sup> s<sup>-1</sup>).  At the daily time resolution it is reported as mean W m<sup>-2</sup>. Positive values correspond to a transfer of water vapor from the ecosystem to the atmosphere.

**Motivation**

Latent heat measures the water loss from an ecosystem to the atmosphere through evapotranspiration (transpiration through plants + evaporation from surfaces).  

Forecasting latent heat (evapotranspiration) can provide insights to water stress for plants and the efficiency that plants are using water relative to NEE, and to the amount of liquid water remaining in the soil for soil moisture forecasting

### Focal sites

Information on the sites can be found here:

```{r message = FALSE}
site_data <- readr::read_csv("https://raw.githubusercontent.com/eco4cast/neon4cast-terrestrial/master/Terrestrial_NEON_Field_Site_Metadata_20210928.csv")
```

```{r echo = FALSE}
site_data %>% 
  select(field_site_id, field_site_name, field_dominant_nlcd_classes, field_latitude, field_longitude, neon_url) %>% 
  rename(siteID = field_site_id,
         `site name` = field_site_name,
         `vegetation type` = field_dominant_nlcd_classes,
         `latitude` = field_latitude,
         `longtitude` =  field_longitude,
         `NEON site URL` = neon_url) %>% 
  arrange(siteID) %>% 
  knitr::kable()
```

### 30-minute target data calulation

To create the data for evaluation (and training) for NEE and LE we extract NEE and LE that pass the turbulence quality control flags (`qfqm.fluxCo2.turb.qfFinl` = 0 ) provided by NEON and has flux values between -50 and 50 umol CO2 m<sup>-2</sup> s<sup>-1</sup>.

The table with the half-hour NEE, LE, and VSWC has the following columns

- `siteID`: NEON site code (e.g., BART)     
- `time`: YYYY-MM-DD HH:MM for the start of the 30-minute period in UTC   
- `nee`:  umol CO2 m<sup>-2</sup> s<sup>-1</sup>   
- `le`: W m<sup>-2</sup>   
- `nee_sd_intercept`: intercept in the nee observation uncertainty standard deviation   
- `nee_sd_slopeP`: slope in the relationship between nee and observation uncertainty standard deviation for positive values of nee   
- `nee_sd_slopeN`: slope in the relationship between nee and observation uncertainty standard deviation for negative values of nee   
- `le_sd_intercept`: intercept in the le observation uncertainty standard deviation   
- `le_sd_slopeP`:slope in the relationship between le and observation uncertainty standard deviation for positive values of le  
- `le_sd_slopeN`: slope in the relationship between le and observation uncertainty standard deviation for negative values of le

The observation uncertainty estimates for nee and le are derived from the PECAN project and can be used by `sd_nee <- nee_sd_intercept + nee_sd_slopeP * nee`.  They are not supplied by NEON.

Here is the download link and format of the `terrestrial_30min` target file

```{r message = FALSE}
readr::read_csv("https://data.ecoforecast.org/targets/terrestrial_30min/terrestrial_30min-targets.csv.gz", guess_max = 1e6)
```

The code used to generate the targets from NEON data can be found [here](https://github.com/eco4cast/neon4cast-terrestrial/blob/master/02_terrestrial_targets.R)

### Daily target data calulation

To evaluate the models that produce daily flux forecasts, we select only days with at least 24 of 48 half hours that pass the quality control flags .  For these days, we average the half-hours and convert to carbon to daily units (gC/m2/day).  The daily data table has the following columns.

- `siteID`: NEON site code (e.g., BART)  
- `time`: YYYY-MM-DD (the day is determined using UTC time)  
- `nee`: g C m<sup>-2</sup> day<sup>-1</sup>  
- `le`:  W m<sup>-2</sup>

Here is the download link and format of the `terrestrial_daily` target file

```{r message = FALSE}
readr::read_csv("https://data.ecoforecast.org/targets/terrestrial_daily/terrestrial_daily-targets.csv.gz", guess_max = 1e6)
```

The code used to generate the targets from NEON data can be found [here](https://github.com/eco4cast/neon4cast-terrestrial/blob/master/02_terrestrial_targets.R)

## Timeline

Forecasts for a minimum of 35 days can be submitted daily by 11:59 pm UTC throughout 2022. A minimum of 35 days in the future must be forecasted for each submission. For example, a forecast submitted on February 1 should be for at least February 1st – March 7th. New forecasts can be submitted daily as new weather forecasts and observations (e.g., new flux data is released by NEON) become available. The key is that submissions are predictions of the future.

Even that daily submissions are allowed and encouraged as new observations and weather forecasts become available, the automation of forecast generation may be ideal.  There are many ways to automate scripts that are written to download observations and meteorology drivers, generate forecasts, and submit forecasts. Two tools that many have used are cron jobs (see the R package [cronR](https://cran.r-project.org/web/packages/cronR/index.html)){target="_blank"} that execute tasks at user specified times and [github actions](https://youtu.be/dMrUlXi4_Bo){target="_blank"}.  See more at [Frequently Asked Questions]    

Cron jobs work on unix and mac systems.  An example of a script that executes a cron job using R can be found [here](https://github.com/eco4cast/neon4cast-terrestrial/blob/master/cron_automation.R){target="_blank"}.

## Flux data latency

NEON data officially releases the flux data on their data portal and API in monthly data packages.  Data for a given month is scheduled to be released around the 15th of the following month. Once the NEON data for a previous month is released.

NEON is also processing flux data with only a 5 day delay (latency).  Any data that has been processed but not included in a released monthly package is avialable on NEON s3 storage.  The list of files that can be downloaded can found [here](https://s3.data.neonscience.org/neon-sae-files/ods/sae_files_unpublished/sae_file_url_unpublished.csv).  

Our targets file is the combination of NEON's monthly releases and the files on the s3 bucket.  As a result, flux data within 5-days of the restart of a forecast are avaialable to inform the forecast.

The reduction of the latency from monthly to 5-days allows this theme to forecast in real-time - a major advancement for the forecastinc challenge.  Thank you NEON!

## Submissions

Instructions for submitting forecasts are found here: [Submission Instructions]

The specific file format for the terrestrial_daily theme is here: [Terrestrial daily]

The specific file format for the terrestrial_30min theme is here: [Terrestrial 30 min]

## Meterological inputs for modeling

Information about forecasted meteorology that is available for you to use when generating your forecasts can be found here: [Meteorology Inputs] 

## Useful functions

Functions for validating, evaluating and submitting forecasts can be found here [Helpful Functions]

Functions for downloading and working with the meteorology forecasts can be be found here: [Access EFI snapshots of NOAA forecasts at NEON sites]

## Null models

Two null models are automatically generated each day - these are simple baseline models.  The persistence null model uses the most recent measurement of nee or le and predicts that the values will be constant in the future.  The climatology null model futures that the nee or lee will be equal to the historical mean of that day of year.  We apply both the persistence and climatology model to the daily fluxes and the climatology to the 30 minute fluxes

Code for the daily persistence null model can be found [here](https://github.com/eco4cast/neon4cast-terrestrial/blob/master/03_terrestrial_flux_30min_persistance.R){target="_blank}

Code for the daily climatology null model can be found [here](https://github.com/eco4cast/neon4cast-terrestrial/blob/master/03_terrestrial_flux_daily_climatology.R){target="_blank"}

Code for the 30 minutate climatology null model can be found [here](https://github.com/eco4cast/neon4cast-terrestrial/blob/master/03_terrestrial_flux_30min_clim.R){target="_blank"}

## FAQ

Answers to frequency asks questions can be found here: [Frequently Asked Questions]

## Design team 

Pending

## Partners

Data used in the challenge are from the National Ecological Observatory Network ([NEON](https://www.neonscience.org/){target="_blank"})

[Ameriflux](https://ameriflux.lbl.gov/){target="_blank"} is an excellent database of eddy-covariance data, including historical data for some of the four challenge sites.

Terrestrial Ecosystem Research Network [(TERN)](https://www.tern.org.au/){target="_blank"} has been involved in the design of the challenge.

