# Participation

## Participation guidance

### How to participate

Participation requires that teams: 

1. Complete a REGISTRATION for each forecast theme you are participating in and each model you are contributing within a theme
2. Agree to the participation agreement below
3. Submit forecast netCDF or csv file(s) 
4. Provide the metadata xml file documenting the forecast 

One contact person should register on behalf of their team. That contact person will be asked to provide the group members' names, emails, and affiliations so that everyone in the group can receive an invitation to join the Challenge theme Slack channel and access group resources. Teams are allowed and encouraged to join the challenge after the start date of each Challenge theme because there are multiple deadlines to submit forecasts. However, only forecasts submitted by each submission deadline will be officially scored.


### Teams

Teams can be individuals or groups. They can represent institutions or organizations. You will have 25 characters for a team name (e.g., “EFI Null Model”) and 10 characters for the team name ID (no spaces allowed; e.g., `EFI_Null`). 

The registration includes team categories (e.g., undergraduate only, graduate only, multi-institution, etc). Please check all that apply.

If your team wants to submit multiple forecasts, please register a team for each model as only one forecast model per cycle per team is allowed. If there are two different time steps in a challenge theme (e.g., the terrestrial carbon flux theme has a 30-minute and 1-day option), register each as separate teams.

### Slack and GitHub Communication

We strongly encourage participants to use the Challenge theme Slack channels to ask questions, discuss ideas and challenges, and share resources. Overall, we strongly encourage a collegial approach to the Challenge -- this is a friendly competition to move the field forward and bring more people into the community, not a cutthroat competition to win by denying other teams useful information.

GitHub repositories for each Challenge theme will be available with helper code and an example workflow (null models). We encourage teams to contribute code to these repositories (via Pull Request) if they develop additional helper code. This is especially important if an individual or group is going to add additional data constraints to their forecast. Remember, the use of data external to NEON is allowed and encouraged so long as it is publicly available and other teams are notified about it. Also, while most anything could be used to calibrate parameters and constrain initial conditions, only other forecasts (e.g. weather) can be used as drivers/covariates during the actual forecast period. 

## Submission instructions

### Forecast format

Teams will submit their forecasts as a single netCDF or csv file with the following naming convention:

`theme_name-year-month-day-team_name_ID.csv`

or

`theme_name-year-month-day-team_name_ID.nc`

Where `year`, `month`, and `day` are the year, month, and day for the first day of the submitted forecast and the `team_name_ID` is the code for the team name that is specified in the registration (`team_name_ID` is a 10 character name with no spaces in it). 

The `theme_name` options are: `terrestrial_daily`, `terrestrial_30min`, `aquatics`, `beetles`, `ticks`, or `phenology`. 

Forecast netCDF or csv files should have the following columns (csv) or variables (netcdf) that correspond to the columns.

#### Terrestrial

- `time`: YYYY-MM-DD HH:MM UTC of the start of the 30-minute value or YYYY-MM-DD for daily forecasts
- `siteID`: NEON code for site
- `ensemble`*: integer value for forecast replicate within the year and month (i.e. ensemble member or MCMC sample)
- `forecast`: set as 1 for each row (1 = variables were forecasted; a 0 would designate a hindcast which does not apply to submissions to the challenge)
- `data_assimilation`: set as 0 for each row (0 = no data assimilation occurred because it is a forecast)
- `nee`: net ecosystem exchange (umol CO2 m-2 s-1)
- `le`: latent heat (W m-2)
- `vswc`: volumetric soil water content (%)

#### Beetles

- `time`: YYYY-MM-DD of forecast (where the DD is the first day of the week that is forecasted)
- `siteID`: NEON code for site
- `ensemble`*: integer value for forecast replicate within the year and month (i.e. ensemble member or MCMC sample)
- `forecast`: set as 1 for each row (1 = variables were forecasted; a 0 would designate a hindcast which does not apply to submissions to the challenge)
- `data_assimilation`: set as 0 for each row (0 = no data assimilation occurred because it is a forecast)
- `abundance`: abundance of beetles
- `richness`: species richness of beetles 

#### Aquatics

- `time`: YYYY-MM-DD of forecast
- `siteID`: NEON code for site
- `ensemble`*: integer value for forecast replicate within the year and month (i.e. ensemble member or MCMC sample)
- `forecast`: set as 1 for each row (1 = variables were forecasted; a 0 would designate a hindcast which does not apply to submissions to the challenge)
- `data_assimilation`: set as 0 for each row (0 = no data assimilation occurred because it is a forecast)
- `oxygen`: dissolved oxygen (ug/L)
- `temp`: water temperature (C)

Here is an example of a forecast file that meets the standard for the aquatics theme

```{r}
readr::read_csv("https://data.ecoforecast.org/forecasts/aquatics/aquatics-2021-03-01-EFInull.csv.gz")
```

#### Phenology

- `time`: YYYY-MM-DD
- `siteID`: NEON code for site
- `ensemble`*: integer value for forecast replicate within the year and month (i.e. ensemble member or MCMC sample)
- `forecast`: set as 1 for each row (1 = variables were forecasted; a 0 would designate a hindcast which does not apply to submissions to the challenge)
- `data_assimilation`: set as 0 for each row (0 = no data assimilation occurred because it is a forecast)
- `gcc_90`: green chromatic coordinate

Here is an example of a forecast file that meets the standard for the phenology theme



#### Ticks

- `time`: YYYY-MM-DD of forecast (where the DD is the first day of the week that is forecasted)
- `siteID`: NEON code for site
- `plotID`: NEON plotID
- `ensemble`*: integer value for forecast replicate within the year and month (i.e. ensemble member or MCMC sample)
- `forecast`: set as 1 for each row (1 = variables were forecasted; a 0 would designate a hindcast which does not apply to submissions to the challenge)
- `data_assimilation`: set as 0 for each row (0 = no data assimilation occurred because it is a forecast)
- `amblyomma_americanum`: Number of *Amblyomma americanum* nymphs per plot per week
- `ixodes_scapularis`: Number of *Ixodes scapularis* nymphs per plot per week

*Teams that are NOT using ensemble-based forecast methods should replace the ensemble column with a statistic column. Multiple statistics can be reported using a long format in a csv or adding a statistic dimension in netCDF. The required options for this column are `mean` and `sd` (standard deviation). You can also include `Conf_interv_02.5`, `Conf_interv_97.5`, `Pred_interval_02.5`, and `Pred_interval_97.5` to describe uncertainty but these are optional. The numbers in the last four options indicate equal-tail quantiles for a 95% interval estimate and `Conf_`=confidence and `Pred_`=predictive. If statistics are reported we will make a Gaussian assumption when calculating error scores. The Continuous Ranked Probability Score is based on the predictive distribution so reported `sd` should be for the predictive distribution.

For those using netCDF, the order of dimensions on forecast variables should be: `time`, `site`, `plot` [ticks only], and `ensemble`.  In practice using a netCDF for the forecast forecast with the `statistic` rather than `ensemble` is not ideal.

Additional detail about file formats can be found in the EFI Forecast Standard Documentation.

### Metadata format

Each submission requires a metadata file to be submitted. The metadata file must have the same name as the submitted forecast, but with the .xml extension. Model descriptions can be uploaded to https://data.ecoforecast.org/minio/submissions/

`theme_name-year-month-day-team_name_ID.xml`

The metadata standard has been designed by the Ecological Forecasting Initiative and is build off the widely used Ecological Metadata Language (EML).

The following components are required:
- Title  
- pubDate: date that forecast is generated    
- License: See below for details  
- Creator(s)  
- Spatial and temporal coverage   
- File descriptions  
- Model timestep   
- forecast_horizon (length)   
- forecast_issue_time   
- forecast_iteration_id   
- forecast_project_id: team_name_ID (e.g., `EFI_NULL`)    
- model_description which includes: forecast_model_id, model name, type (statistical, process-based, machine-learning, etc), and repository (url or DOI).   
- Info on model structure and uncertainty (standards section 2.1.2)    

The license for the forecast output is required to be from the following Creative Commons License options: CC BY, CC BY-SA, CC BY-NC, CC BY-NC-SA. While we recommend a CC BY license, teams may use less permissive CC licenses if more appropriate. The license entry can be the CC option (i.e., CC BY) and a web link to the full CC license (e.g., https://creativecommons.org/licenses/by/4.0/)

We recommend teams read the full metadata standard description for definitions and more information, and in particular that they look at the example vignettes, which demonstrate the standard being used. Note that these Standards are a work in progress. If you find issues as you are applying them, let us know at eco4cast.initaitive@gmail.com.

The Ecological Forecasting Initiative has provided R scripts to assist in generating the metadata XML file. The scripts can be found at the GitHub repository for the standard: https://github.com/eco4cast/EFIstandards as well as the EML validator. Teams are encouraged to check the validity of their metadata before submission.

### Submission process

Individual forecast (csv, netCDF) and metadate (xml) files can be uploaded any time before the specific deadlines as defined by each theme. Only the most recent files will be scored. 

Teams will submit their forecast netCDF or csv files through the challenge website. You can manually submit your forecast through the https://data.ecoforecast.org/minio/submissions/ website using the red plus on the bottom left. You can submit from an R script using the following:

```{r eval = FALSE}
Sys.setenv("AWS_DEFAULT_REGION" = "data",
           "AWS_S3_ENDPOINT" = "ecoforecast.org")

aws.s3::put_object(object = "theme_name-forecast-year-month-day-team_name.csv", bucket = "submissions")
```


Submissions need to adhere to the forecast format that is provided above, including the file naming convention. Our cyberinfastructure automatically evaluates forecasts and relies on the expected formatting. Contact eco4cast.initiative@gmail.com if you experience technical issues with submitting.

Note: If you have used AWS in the past you might have credential files in an .aws folder in your home directory that will cause an error when you try to upload to a non-AWS bucket. If you encounter this error you may need to rename your credentials files so put_object doesn't try to read them.

### Archiving models

Teams are highly encouraged to publicly archive the code they are using for their forecast models and workflows. Information about where models are archived would be included in your metadata XML.

Teams are also encouraged to use Docker or Singularity to containerize their models & workflows. EFI conventions for containerizing forecasts are still being developed, but our aim (particularly in later years of the forecast challenge) is to be able to provide shared cyberinfrastructure that makes it easier for teams to automate containerized forecasts. Containers will also facilitate Challenge themes interested in performing post-hoc analyses, such as uncertainty quantification and sensitivity analysis.

### Computational reesources
We are currently working with CyVerse for access to computational resources for teams that require resources not available through home institutions. We will update with more details as they become available.

## Participation agreement

All participants agree to have their forecast posted in real-time on the NEON Ecological Forecast Challenge Output RShiny app (in development) and potentially published in a scientific journal. The manuscripts describing the accuracy of forecasts across teams will be coordinated by the Ecological Forecasting Initiative Research Coordination Network and extend authorship to members of each team with an opt-in policy. 

If a publication is generated by a forecast team, we ask that the manuscript acknowledge the Ecological Forecasting Initiative Research Coordination Network and its support from the National Science Foundation (DEB-1926388).

## NEON Data Use

NEON data products, software, and derivatives thereof are freely available for use when accompanied by appropriate disclaimers, acknowledgments, and data citations, defined in the NEON data use policy.

## Additional data options

Individuals and groups may create forecasts that use other publicly available data in addition to the NEON data, so long as other teams participating in the challenge are notified about the existence of the data via the Challenge theme’s Slack channel. Teams are encouraged to make available the code they are using to access, download, and process any additional data constraints they are using, ideally via a pull request to each Challenge Github repo. 

When considering the use of data in forecasts it is important to distinguish data that are being used as drivers/covariates during each forecast from data being used to constrain model structure, parameters, initial conditions, and error distributions. While the latency of NEON data requires that some of our forecast will be (fully or partly) hindcasts, all forecasts should be run as if they are true forecasts -- you cannot use any observed data as a driver/covariate or constraint during the forecast period itself as that info would not have been available at the forecast start date. For example, if you find that a particular variable is a useful covariate during the model development & calibration period (e.g. soil temperature) then you would need to find or make a forecast of that variable if you want to use it as a covariate. Teams using meteorological covariates should use the shared meteorological driver data provided by EFI (see Shared Forecast Drivers).

As an example of potentially useful external data, each NEON site has subsets of various remote sensing products that are hosted on the ORNL DAAC (ORNL DAAC subsets). These include:

MODIS collection 6: LAI, FPAR, burned area, surface reflectance, land surface temperature, vegetation indices (NDVI, EVI), modeled ET, GPP, NPP.

VIIRS collection 1: surface reflectance, vegetation indices, LAI, FPAR, land surface temperature,

SMAP: modeled NEE, GPP, Rh, SOC

Daymet: daily surface weather data
